Chaotic mixing of tidal debris

### 2015-06-09 ms.tex
change to superfreq

This potential is a simple and unrealistic model for the total potential of a Milky-Way-like galaxy, however it represents a conservative choice for exploring the structure of orbits in the halos of such galaxies. Realistic galactic potentials will have a significant component due to the disk and bulge, may have twisting inertia tensors \citep{romanowsky98}, radially changing axis ratios \citep[e.g.,][]{kazantzidis04,debattista08,veraciro11,butsky15}, significant substructure \citep{moore98,zemp09}, or time dependence \citep[either from bulk rotation, mass growth, mergers, etc.; see, e.g.,][]{bailin05}. We expect inclusion of any of these effects to increase the amount and influence of chaos; see the Discussion (Section~\ref{sec:discussion}) for a few simple demonstrations.

@@

This method is referred to as ``Numerical Approximation of Fundamental Frequencies'' (NAFF) and has been used extensively in planetary dynamics \citep[e.g.,][]{laskar93b, laskar96} and galaxy dynamics \citep{papaphilippou98, valluri98}, especially in the study of orbits in triaxial systems. We have implemented and tested our own open-source implementation of NAFF in \texttt{Python}.\footnote{This code is publicly available through the \texttt{Gary} package, \url{http://...}} Though our implementation differs slightly from the original definition and from that used in \cite{valluri98}, we still refer to the method as NAFF. For more details about NAFF, see Appendix~\ref{sec:naffapdx} or, e.g., \cite{laskar88, laskar93, papaphilippou96}.

If an orbit is chaotic, the motion can no longer be expanded in terms of a single set of fundamental frequencies. For a weakly chaotic orbit, the orbit may appear consistently periodic over long windows of time. NAFF will pick out a set of frequencies for such orbits that correspond to the largest peaks in the power spectrum of the stochastic orbits, however these peaks will change location and amplitude with time. For more strongly chaotic orbits, the power spectrum will be quite noisy and the peak frequencies may change erratically when comparing two separate sections of orbit. The frequencies picked out by NAFF for such orbits will therefore represent the average periodic nature of the orbit over a given period. We define the fractional frequency diffusion rate per orbital period, $\mathcal{R}$, in the $i$th fundamental frequency as
\begin{equation}
	\mathcal{R}_i = \frac{\Omega_i^{(2)} - \Omega_{i}^{(1)}}{\Omega_{i}^{(1)} \, N} \label{eq:fdrate}
\end{equation}
where $\Omega_i^{(1)}$ and $\Omega_i^{(2)}$ is the $i$th fundamental frequency computed in each of two consecutive sections of orbit and $N$ is the number of orbital periods, ${\rm T}$, in a single section of orbit. From this rate, we can compute a timescale; for a small ensemble of orbits (e.g., tidal debris), the relevant timescale is the time over which the change in frequencies for a single orbit is comparable to the spread of frequencies in the ensemble. We therefore define the frequency diffusion time (in units of number of orbital periods) with an unspecified parameter, $f$,
\begin{equation}
	t_\Omega(f) = f \, (\max_{a_i} \, \mathcal{R}_i)^{-1} \label{eq:fdtime}
\end{equation}
where the maximum is taken with respect to the corresponding amplitudes, $a_i$, of the fundamental frequency components \citep[see][]{valluri12}. When $f = 1$, this is the number of orbital periods over which we expect order-unity changes to the frequencies, whereas $f = 0.01$ corresponds to $1\%$ changes.

@@

For each orbit, we also separately integrate for $\approx$80 orbital periods and use \superfreq\ to compute the fundamental frequencies for the two consecutive sections of 40 orbital periods. We have found that this is the shortest integration window for which we are able to consistently and accurately recover the frequencies using \superfreq.

 Figure~\ref{fig:freqdiff} shows the same grid of initial conditions, but now the greyscale intensity is set by the logarithm of the frequency diffusion time. The darker pixels have shorter frequency diffusion times and are more chaotic. This map reveals the intersection of this particular energy hypersurface with the rich structure of resonant surfaces present in this potential and highlights the accuracy of NAFF --- weak chaos is detectable over just 10's of orbital periods, compared to the many thousands of orbits it would take to detect such features with the maximum Lyapunov exponent. The tube orbits in this potential are mostly regular or only mildly chaotic --- the largest regular regions are associated with the short-axis and long-axis tube orbits --- however islands of stronger chaos do appear, especially at the intersections of resonances where resonance overlap occurs.

The strongest chaotic regions appear in both of the above grids (Figure~\ref{fig:lyapmap} and Figure~\ref{fig:freqdiff}). The resonant structure is revealed in the frequency diffusion map from integrations of just 100 orbital periods, though the majority of the orbits have estimated chaotic timescales much longer than 100 orbital periods. In the next section, we analyze the density evolution of finite-volume ensembles of orbits around each orbit in the above grids and compare the density evolution of the ensembles to the single-orbit chaos indicators computed in this section.

@@

The Lyapunov time and frequency diffusion rate measure the timescales over which chaos is relevant \emph{local} to a given orbit --- that is, these quantities are measures of how infinitesimal deviations will diverge on average from some parent orbit of interest, or of how long it takes for the frequencies to change by some amount. Tidal debris is disrupted from progenitor systems with finite spreads in orbital properties (e.g., energy). For a disrupting, globular-cluster-scale progenitor, the typical energy dispersion of the debris is 0.1--1\% of the progenitor orbital energy, but for a dwarf-galaxy-scale progenitor, the dispersion can be 5--20\%. Does the Lyapunov or frequency diffusion time predict the timescale over which a finite phase-space volume (e.g., tidal debris) stays coherent?

We start by considering three particular orbits chosen from the orbit grid of Section~\ref{sec:results1}: a regular orbit (R), a weakly chaotic orbit (W), and a strongly chaotic orbit (S). The orbits were chosen at random from a single column of the orbit grid based on their frequency diffusion time, and the initial conditions and chaos diagnostics are listed in Table~\ref{tbl:three-orbits}. We use the Self-Consistent Field (SCF) basis function expansion code \citep{hernquist92} to run N-body simulations of globular-cluster-mass progenitor systems on these three orbits. The initial mass of the clusters is \todo{kvj}{2.5E4 \msun contained within what again? 35 scale radii? but this is plummer??}. For each simulation, the progenitor is initialized as a $10^4$ particle Plummer sphere and the background triaxial NFW potential is turned on slowly over 250 Myr to reduce artificial gravitational shocking. We start the progenitor systems at apocenter and evolve for $\approx$16 orbital periods so that each simulation finishes again at subsequent apocenter. Figure~\ref{fig:nbodysims} shows the final particle distributions rotated so that the angular momentum of the progenitor orbit is aligned with the $z$-axis. A thin stream forms on the regular orbit (left column), a more diffuse stream on the mildly chaotic orbit (middle column), and a ``fanned'' stream on the strongly chaotic orbit (right column). Given the long Lyapunov and frequency diffusion times of the mildly chaotic orbit, it is surprising that stream formed on this orbit appears to be more diffuse than the regular orbit stream.

@@

Both the Lyapunov exponent and the frequency diffusion rate measure mean, long-term rates of chaotic diffusion. If a weakly chaotic orbit is confined (by other nearby, non-overlapping resonances or stable regions), the mean drift of an orbit in frequency space may be small if computed over timescales long compared to the orbital time but short compared to the Arnold diffusion time. For the potential considered in this paper, it turns out that these are comparable to the timescales for which the frequency diffusion rate of the previous section were computed ($\approx$100-1000 orbital periods). This window length was chosen in order to have a strong enough signal in the frequency spectrum of chaotic orbits to measure the most significant frequencies using NAFF, however, it erases the small-scale variation of the frequencies.

As a demonstration of the small-scale frequency modulation, we consider again the three orbits of Section~\ref{sec:results2} (initial conditions are listed in Table~\ref{tbl:three-orbits}). Figure~\ref{fig:fdiff-zooms} shows zoomed-reruns of the frequency diffusion time map of Figure~\ref{fig:freqdiff} centered on the three orbits --- the pixel positions are now shown relative to the three parent orbits. The frequency diffusion times shown in these grids were computed with two consecutive windows of length 35 orbital periods (shorter than the windows used in the previous, large grid of Figure~\ref{fig:freqdiff}). The resonant structure around the mildly chaotic orbit (white regions in panel W) bound the orbit to a small region, whereas for the strongly chaotic orbit (S), the entire region surrounding the orbit is strongly chaotic. To resolve the short-time behavior of the frequency diffusion (corresponding to motion across resonance layers) we compute the frequencies in a series of rolling windows along each orbit. We use a window with a width equal to 35 orbital periods and shift the window by half an orbital period between each calculation of the most significant frequencies --- this is the smallest window size we could choose before NAFF begins to consistently fail to recover the frequencies for the strongly chaotic orbit.  Figure~\ref{fig:three-orbits-freqs} shows plots of the frequency ratios as computed in each window of time --- plotted are the percent deviations of the frequency ratios from the median value over all windows. The difference between the first window (green +) and the last window (red x) is 50 orbital periods. The frequencies of the weakly chaotic orbit (panel labeled $W$) evolve quickly but are bounded to a small volume with fractional size $\approx$0.3\% (presumably by nearby resonant surfaces). This is comparable to the frequency spread in globular cluster tidal debris (0.1--0.5\%).

>>>>

This potential is a simple and unrealistic model for the total potential of a Milky-Way-like galaxy, however it represents a conservative choice for exploring the structure of orbits in the halos of such galaxies. Realistic galactic potentials will have a significant component due to the disk and bulge, may have twisting inertia tensors \citep{romanowsky98}, radially changing axis ratios \citep[e.g.,][]{kazantzidis04,debattista08,veraciro11,butsky15}, significant substructure \citep{moore98,zemp09}, or time dependence \citep[either from bulk rotation, mass growth, mergers, etc.; see, e.g.,][]{bailin05}. We expect inclusion of any of these effects to increase the amount and influence of chaos; see the Discussion (Section~\ref{sec:discussion}) for a few simple demonstrations. \todo{apw}{Cite this paper: hunter05 for disk crossings and chaos}

@@

This method is referred to as ``Numerical Approximation of Fundamental Frequencies'' (NAFF) and has been used extensively in planetary dynamics \citep[e.g.,][]{laskar93b, laskar96} and galaxy dynamics \citep{papaphilippou98, valluri98}, especially in the study of orbits in triaxial systems. We have implemented and tested a version of this procedure in the \project{Python} programming language. Our implementation differs slightly from the original definition and from that used in \cite{valluri98}: we have found that using a higher order window function \citep[e.g.,][]{hunter02} of the form
\begin{equation}
	W(t/\inttime) = \frac{2^p \, (p!)^2}{(2p)!} \left( 1 + \cos \frac{\pi t}{\inttime}\right)^p
\end{equation}
with $p=2$ allows for more reliable determination of frequencies from strongly chaotic orbits. We refer to this slight modification of the algorithm as \superfreq\ and the code is open-source and publicly available on \project{GitHub} at \url{https://github.com/adrn/SuperFreq}. For more details about the algorithm, see Appendix~\ref{sec:naffapdx} or \cite{laskar88, laskar93, papaphilippou96}.

If an orbit is chaotic, the motion can no longer be expanded in terms of a single set of fundamental frequencies. For a weakly chaotic orbit, the orbit may appear consistently periodic over long windows of time. \superfreq\ will pick out a set of frequencies for such orbits that correspond to the largest peaks in the power spectrum of the stochastic orbits, however these peaks will change location and amplitude with time. For more strongly chaotic orbits, the power spectrum will be quite noisy and the peak frequencies may change erratically when comparing two separate sections of orbit. The frequencies picked out by \superfreq\ for such orbits will therefore represent the average periodic nature of the orbit over a given period. We define the fractional frequency diffusion rate per orbital period, $\mathcal{R}$, in the $i$th fundamental frequency as
\begin{equation}
	\mathcal{R}_i = \frac{\Omega_i^{(2)} - \Omega_{i}^{(1)}}{\Omega_{i}^{(1)} \, N} \label{eq:fdrate}
\end{equation}
where $\Omega_i^{(1)}$ and $\Omega_i^{(2)}$ is the $i$th fundamental frequency computed in each of two consecutive sections of orbit and $N$ is the number of orbital periods in a single section of orbit. By inverting this rate, we can compute the timescale over which we expect order-unity changes to the fundamental frequencies: the \emph{frequency diffusion time} is defined as
\begin{equation}
	t_\Omega = \, (\max_{a_i} \, \mathcal{R}_i)^{-1} \label{eq:fdtime}
\end{equation}
where the maximum is taken with respect to the corresponding amplitudes, $a_i$, of the fundamental frequency components \citep[see][]{valluri12}, and the time is in units of number of orbital periods, ${\rm T}$. For a small ensemble of orbits (e.g., tidal debris), the relevant timescale is the time over which the change in frequencies for a single orbit is comparable to the spread of frequencies in the ensemble. We can estimate this timescale by multiplying the frequency diffusion time by a factor equal to the fractional spread in frequencies of the debris. For example, a globular cluster typically has $\approx$1\% spreads in fundamental frequencies, so by multiplying the frequency diffusion time by $f = 0.01$, we can estimate the time (number of orbital periods) over which we expect the frequencies to evolve by this amount.

@@

For each orbit, we also separately integrate for $\approx$80 orbital periods and use \superfreq\ to compute the fundamental frequencies for the two consecutive sections of 40 orbital periods. We have found that this is the shortest integration window for which we are able to consistently and accurately recover the frequencies using \superfreq. Figure~\ref{fig:freqdiff} shows the same grid of initial conditions, but now the greyscale intensity is set by the logarithm of the (order-unity) frequency diffusion time. The darker pixels have shorter frequency diffusion times and are more chaotic. This map reveals the intersection of this particular energy hypersurface with the rich structure of resonant surfaces present in this potential and highlights the accuracy of \superfreq\ --- weak chaos (grey) is detectable over just 10's of orbital periods, compared to the many thousands of orbits it would take to detect such features with the maximum Lyapunov exponent. The tube orbits in this potential are mostly regular or only mildly chaotic --- the largest regular regions are associated with the short-axis and long-axis tube orbits --- however islands of stronger chaos do appear, especially at the intersections of resonances where resonance overlap occurs.

The strongest chaotic regions appear in both of the above grids (Figure~\ref{fig:lyapmap} and Figure~\ref{fig:freqdiff}). The resonant structure is revealed in the frequency diffusion map from integrations of just 80 orbital periods, though the majority of the orbits have estimated chaotic timescales corresponding to thousands of orbital periods. In the next section, we analyze the density evolution of finite-volume ensembles of orbits around each orbit in the above grids and compare the density evolution of the ensembles to the single-orbit chaos indicators computed in this section.

@@

The Lyapunov time and frequency diffusion rate measure the timescales over which chaos is relevant \emph{local} to a given orbit --- that is, these quantities are measures of how infinitesimal deviations will diverge on average from some parent orbit of interest, or of how long it takes for the frequencies to change by some amount. Tidal debris is disrupted from progenitor systems with finite spreads in orbital properties (e.g., energy). For a disrupting, globular-cluster-scale progenitor, the typical energy dispersion of the debris is 0.1--1\% of the progenitor orbital energy \citep[assuming masses of $10^4$--$10^5$~\msun;][]{johnston98}, but for a dwarf-galaxy-scale progenitor, the dispersion can be $\gtrsim$10\%. Does the Lyapunov or frequency diffusion time predict the timescale over which a finite phase-space volume (e.g., tidal debris) stays coherent?

We start by considering three particular orbits chosen from the orbit grid of Section~\ref{sec:results1}: a regular orbit (R), a weakly chaotic orbit (W), and a strongly chaotic orbit (S). The orbits were chosen at random from a single column of the orbit grid based on their frequency diffusion time, and the initial conditions and chaos diagnostics are listed in Table~\ref{tbl:three-orbits}. We use the Self-Consistent Field (SCF) basis function expansion code \citep{hernquist92} to run N-body simulations of globular-cluster-mass progenitor systems on these three orbits. The initial mass of the clusters is \todo{kvj}{2.5E4 \msun contained within what again? 35 scale radii? but this is plummer??}. For each simulation, the progenitor is initialized as a $10^4$ particle Plummer sphere and the background triaxial NFW potential is turned on slowly over 250 Myr to reduce artificial gravitational shocking. \todo{apw}{Label the ICs for the 3 orbits on Figures 3 and 4} We start the progenitor systems at apocenter and evolve for $\approx$16 orbital periods so that each simulation finishes again at subsequent apocenter. Figure~\ref{fig:nbodysims} shows the final particle distributions rotated so that the angular momentum of the progenitor orbit is aligned with the $z$-axis. A thin stream forms on the regular orbit (left column), a more diffuse stream on the mildly chaotic orbit (middle column), and a ``fanned'' stream on the strongly chaotic orbit (right column). Given the long Lyapunov and frequency diffusion times of the mildly chaotic orbit, it is surprising that stream formed on this orbit appears to be more diffuse than the regular orbit stream.

@@

Both the Lyapunov exponent and the frequency diffusion rate measure mean, long-term rates of chaotic diffusion. If a weakly chaotic orbit is confined (by other nearby, non-overlapping resonances or stable regions), the mean drift of an orbit in frequency space may be small if computed over timescales long compared to the orbital time but short compared to the Arnold diffusion time. For the potential considered in this paper, it turns out that these are comparable to the timescales for which the frequency diffusion rate of the previous section were computed ($\approx$100-1000 orbital periods). This window length was chosen in order to have a strong enough signal in the frequency spectrum of chaotic orbits to measure the most significant frequencies using \superfreq\, however, it erases the small-scale variation of the frequencies.

As a demonstration of the small-scale frequency modulation, we consider again the three orbits of Section~\ref{sec:results2} (initial conditions are listed in Table~\ref{tbl:three-orbits}). Figure~\ref{fig:fdiff-zooms} shows zoomed-reruns of the frequency diffusion time map of Figure~\ref{fig:freqdiff} centered on the three orbits --- the pixel positions are now shown relative to the three parent orbits. The frequency diffusion times shown in these grids were computed with two consecutive windows of length 35 orbital periods (shorter than the windows used in the previous, large grid of Figure~\ref{fig:freqdiff}). The resonant structure around the mildly chaotic orbit (white regions in panel W) bound the orbit to a small region, whereas for the strongly chaotic orbit (S), the entire region surrounding the orbit is strongly chaotic. To resolve the short-time behavior of the frequency diffusion (corresponding to motion across resonance layers) we compute the frequencies in a series of rolling windows along each orbit. We use a window with a width equal to 35 orbital periods and shift the window by half an orbital period between each calculation of the most significant frequencies --- this is the smallest window size we could choose before \superfreq\ begins to consistently fail to recover the frequencies for the strongly chaotic orbit.  Figure~\ref{fig:three-orbits-freqs} shows plots of the frequency ratios as computed in each window of time --- plotted are the percent deviations of the frequency ratios from the median value over all windows. The difference between the first window (green +) and the last window (red x) is 50 orbital periods. The frequencies of the weakly chaotic orbit (panel labeled $W$) evolve quickly but are bounded to a small volume with fractional size $\approx$0.3\% (presumably by nearby resonant surfaces). This is comparable to the frequency spread in globular cluster tidal debris (0.1--0.5\%).
